---
title: "NA vignette"
output:
  pdf_document:
    toc: yes
    number_sections: yes
    toc_depth: 4
  html_document:
    toc: yes
    keep_md: yes
    toc_depth: 4
  github_document:
    toc: yes
---

```{r setup, include=FALSE}
library(DAAG)
knitr::opts_chunk$set(echo = TRUE)
```

Most of this code came from Harvard STAT 109 class, Prof. Bharatendra Rai. Material used here for educational purposes. It is available in YouTube and GitHub. See links under references. I expanded the material with my own notes and R documentation and I plan to continue adding examples overtime.


# Datasets from DAAG to test NA

```{r}
data(package='DAAG')
```

## Rainforest example

### Looking for NA and missing data

Reference: Dr. Bharantendra https://www.youtube.com/watch?v=Q7gYkpSi8Nk&list=PL34t5iLfZddtUUABMikey6NtL05hPAp42&index=11

```{r}

data('rainforest')
# str(rainforest)
summary(rainforest)
dim(rainforest)
```

Notice the extra row with NA values. You will not be able to calculate the mean or the sd.

### Ignoring NA 

```{r}

mean(rainforest$wood, na.rm = TRUE)

```
## Science example

```{r}
# Load the data
data('science')
summary(science)
dim(science)
```

### Identify rows with NA

Use !complete.cases(dataset) which mean NOT complete cases, those with NA.

```{r}
science[!complete.cases(science),]

```

### Save dataset without NA

```{r}
science_wo_na <- na.omit(science)
summary(science_wo_na)
dim(science_wo_na)

```

