---
title: "Statistical distrivutions vignette"
output:
  html_document:
    toc: yes
---

# Purpose

This vignette aims to introduce you statistical distributions using R.

# Reference

Dr. Bharatendra https://www.youtube.com/watch?v=rsfV57N7Uns&list=PL34t5iLfZddtUUABMikey6NtL05hPAp42&index=7 

# Libraries

```{r setup, include=FALSE}

# Add DAAG to pull some data
library(DAAG)

# This is the library we will highlight here
library(dplyr)

knitr::opts_chunk$set(echo = TRUE)
```


# Normal distribution

## Generate the data

```{r}

z <- pretty(c(-5, 5), 100)

ht <- dnorm(z)

# plot type line
plot(z, ht, type='l')

```

## prob (z < a certain point)

```{r}

# P(z < -2)
pnorm(-2)

```

## Get point where prob = 0.95

At what point will I have 95%

```{r}

# P(z < ?) = 0.95
qnorm(.95)
```


## Now for a diff mean

```{r}

# Where mean is 0.2 and sigma is 0.5
qnorm(.95, 2, 0.5)

```

# t distribution

## Generate data

```{r}
t <- pretty(c(-5, 5), 100)

ht <- dt(t, 3)

# plot type line
plot(t, ht, type='l')


```

## prob (t < a certain point)

With 3 degrees of freedom

```{r}

# P(z < -2)
pt(-2, 3)

```

## Get point where prob = 0.95

At what point will I have 95%, with 3 degrees of freedom

```{r}

# P(z < ?) = 0.95
qt(.95, 3)
```


# Q-Q plot (Normal)

## Default

```{r}
x <- rnorm(10)
qqnorm(x)
```


## Add a line

```{r}
qqnorm(x)
qqline(x)
```


