---
title: "Neural networks on 2D and 3D datasets"
author: "Oscar A. Trevizo"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: yes
    number_sections: yes
    toc_depth: 4
  # html_document:
  #   toc: yes
  #   keep_md: yes
  #   toc_depth: 4
  github_document:
    toc: yes

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

crimson <- '#A51C30'
royalblue <- '#002366'
slate <- '#8996A0'
ivy <- '#52854C'
saffron <- '#D16103'
indigo <- '#293352'
shade <- '#BAC5C6'

# https://data.library.virginia.edu/setting-up-color-palettes-in-r/
cc <- palette()
palette(c(crimson, royalblue, slate, ivy, saffron, indigo, shade))
#colors <- palette(rainbow_hcl(2))
colors <- palette()

```

# Load the libraries

```{r}
library(neuralnet)
library(scatterplot3d)
library(grid)

```

# Functions

## 3D dataset and plotting

```{r}
##
#
# 3D dataset with 2 categories: 0 and 1
# Each observation has three variables x, y, z.
# The function creates 3 columns (x, y, z) normally distributed N(0,1) values.
# Category 0 occurs if all x, y, z data point are within a cube. Otherwise category 1.
#
# Arguments:
#   n: Number of observations (rows)
#   b: Cube boundary (float)
#
#
bld_3d_2c <- function(n = 1000, b = 1.25){

  # generate a matrix with normally distributed values N(0,1)
  xyz <- matrix(rnorm(3*n), ncol=3)
  
  # Create a vector of n observations init to 0, for class 0
  cl <- rep(0, n)

  # Test each xyz term to see if any value falls out of its limit
  # If the value falls out of its limit, assign class 1
  for(i in 1:n){
    if (xyz[i,1] > b) cl[i] = 1
    if (xyz[i,2] > b) cl[i] = 1
    if (xyz[i,3] > b) cl[i] = 1
    if (xyz[i,1] < -b) cl[i] = 1
    if (xyz[i,2] < -b) cl[i] = 1
    if (xyz[i,3] < -b) cl[i] = 1
  }

  # Now bind the columns to return all in one matrix
  xyz <- cbind(cl,xyz)
  
  colnames(xyz) <- c("cl", "x", "y", "z")
  
  # Returns a matrix and then one can change to dataframe downstream
  # xyz <- as.data.frame(xyz)
  # 
  # xyz$cl <- as.factor(xyz$cl)

  return(xyz)
}

## 
#
# From help(scatterplot3D) example 6; by Martin Maechler
#
cubedraw <- function(res3d, min = 0, max = 255, cex = 2, text. = FALSE)
{
  ## Purpose: Draw nice cube with corners
  cube01 <- rbind(c(0,0,1), 0, c(1,0,0), c(1,1,0), 1, c(0,1,1), # < 6 outer
                  c(1,0,1), c(0,1,0)) # <- "inner": fore- & back-ground
  cub <- min + (max-min)* cube01
  ## visibile corners + lines:
  res3d$points3d(cub[c(1:6,1,7,3,7,5) ,], cex = cex, type = 'b', lty = 1)
  ## hidden corner + lines
  res3d$points3d(cub[c(2,8,4,8,6),     ], cex = cex, type = 'b', lty = 3)
  if(text.)## debug
      text(res3d$xyz.convert(cub), labels=1:nrow(cub), col=crimson, cex=2)
}


```

## Nueral Net functions

Credit for this funcitons goes to Harvard statistical learning class by professors Andrey Sivachenko, PhD and Victor A. Farutin, PhD.

# Preface functions

```{r, echo=FALSE}
# From Harvard statistical learning class by professors Andrey Sivachenko, PhD and Victor A. Farutin, PhD.

# I copied these function here for learning purposes. All credit goes to Harvard's data science program.

plot.nn <-
function (x, rep = NULL, x.entry = NULL, x.out = NULL, radius = 0.15, 
    arrow.length = 0.2, intercept = TRUE, intercept.factor = 0.4, 
    information = TRUE, information.pos = 0.1, col.entry.synapse = "black", 
    col.entry = "black", col.hidden = "black", col.hidden.synapse = "black", 
    col.out = "black", col.out.synapse = "black", col.intercept = "blue", 
    fontsize = 12, dimension = 6, show.weights = TRUE, file = NULL, 
    ...) 
{
    net <- x
    if (is.null(net$weights)) 
        stop("weights were not calculated")
    if (!is.null(file) && !is.character(file)) 
        stop("'file' must be a string")
    if (is.null(rep)) {
        for (i in 1:length(net$weights)) {
            if (!is.null(file)) 
                file.rep <- paste(file, ".", i, sep = "")
            else file.rep <- NULL
            #dev.new()
            plot.nn(net, rep = i, x.entry, x.out, radius, arrow.length, 
                intercept, intercept.factor, information, information.pos, 
                col.entry.synapse, col.entry, col.hidden, col.hidden.synapse, 
                col.out, col.out.synapse, col.intercept, fontsize, 
                dimension, show.weights, file.rep, ...)
        }
    }
    else {
        if (is.character(file) && file.exists(file)) 
            stop(sprintf("%s already exists", sQuote(file)))
        result.matrix <- t(net$result.matrix)
        if (rep == "best") 
            rep <- as.integer(which.min(result.matrix[, "error"]))
        if (rep > length(net$weights)) 
            stop("'rep' does not exist")
        weights <- net$weights[[rep]]
        if (is.null(x.entry)) 
            x.entry <- 0.5 - (arrow.length/2) * length(weights)
        if (is.null(x.out)) 
            x.out <- 0.5 + (arrow.length/2) * length(weights)
        width <- max(x.out - x.entry + 0.2, 0.8) * 8
        radius <- radius/dimension
        entry.label <- net$model.list$variables
        out.label <- net$model.list$response
        neuron.count <- array(0, length(weights) + 1)
        neuron.count[1] <- nrow(weights[[1]]) - 1
        neuron.count[2] <- ncol(weights[[1]])
        x.position <- array(0, length(weights) + 1)
        x.position[1] <- x.entry
        x.position[length(weights) + 1] <- x.out
        if (length(weights) > 1) 
            for (i in 2:length(weights)) {
                neuron.count[i + 1] <- ncol(weights[[i]])
                x.position[i] <- x.entry + (i - 1) * (x.out - 
                  x.entry)/length(weights)
            }
        y.step <- 1/(neuron.count + 1)
        y.position <- array(0, length(weights) + 1)
        y.intercept <- 1 - 2 * radius
        information.pos <- min(min(y.step) - 0.1, 0.2)
        if (length(entry.label) != neuron.count[1]) {
            if (length(entry.label) < neuron.count[1]) {
                tmp <- NULL
                for (i in 1:(neuron.count[1] - length(entry.label))) {
                  tmp <- c(tmp, "no name")
                }
                entry.label <- c(entry.label, tmp)
            }
        }
        if (length(out.label) != neuron.count[length(neuron.count)]) {
            if (length(out.label) < neuron.count[length(neuron.count)]) {
                tmp <- NULL
                for (i in 1:(neuron.count[length(neuron.count)] - 
                  length(out.label))) {
                  tmp <- c(tmp, "no name")
                }
                out.label <- c(out.label, tmp)
            }
        }
        grid.newpage()
        pushViewport(viewport(name = "plot.area", width = unit(dimension, 
            "inches"), height = unit(dimension, "inches")))
        for (k in 1:length(weights)) {
            for (i in 1:neuron.count[k]) {
                y.position[k] <- y.position[k] + y.step[k]
                y.tmp <- 0
                for (j in 1:neuron.count[k + 1]) {
                  y.tmp <- y.tmp + y.step[k + 1]
                  result <- calculate.delta(c(x.position[k], 
                    x.position[k + 1]), c(y.position[k], y.tmp), 
                    radius)
                  x <- c(x.position[k], x.position[k + 1] - result[1])
                  y <- c(y.position[k], y.tmp + result[2])
                  grid.lines(x = x, y = y, arrow = arrow(length = unit(0.15, 
                    "cm"), type = "closed"), gp = gpar(fill = col.hidden.synapse, 
                    col = col.hidden.synapse, ...))
                  if (show.weights) 
                    draw.text(label = weights[[k]][neuron.count[k] - 
                      i + 2, neuron.count[k + 1] - j + 1], x = c(x.position[k], 
                      x.position[k + 1]), y = c(y.position[k], 
                      y.tmp), xy.null = 1.25 * result, color = col.hidden.synapse, 
                      fontsize = fontsize - 2, ...)
                }
                if (k == 1) {
                  grid.lines(x = c((x.position[1] - arrow.length), 
                    x.position[1] - radius), y = y.position[k], 
                    arrow = arrow(length = unit(0.15, "cm"), 
                      type = "closed"), gp = gpar(fill = col.entry.synapse, 
                      col = col.entry.synapse, ...))
                  draw.text(label = entry.label[(neuron.count[1] + 
                    1) - i], x = c((x.position - arrow.length), 
                    x.position[1] - radius), y = c(y.position[k], 
                    y.position[k]), xy.null = c(0, 0), color = col.entry.synapse, 
                    fontsize = fontsize, ...)
                  grid.circle(x = x.position[k], y = y.position[k], 
                    r = radius, gp = gpar(fill = "white", col = col.entry, 
                      ...))
                }
                else {
                  grid.circle(x = x.position[k], y = y.position[k], 
                    r = radius, gp = gpar(fill = "white", col = col.hidden, 
                      ...))
                }
            }
        }
        out <- length(neuron.count)
        for (i in 1:neuron.count[out]) {
            y.position[out] <- y.position[out] + y.step[out]
            grid.lines(x = c(x.position[out] + radius, x.position[out] + 
                arrow.length), y = y.position[out], arrow = arrow(length = unit(0.15, 
                "cm"), type = "closed"), gp = gpar(fill = col.out.synapse, 
                col = col.out.synapse, ...))
            draw.text(label = out.label[(neuron.count[out] + 
                1) - i], x = c((x.position[out] + radius), x.position[out] + 
                arrow.length), y = c(y.position[out], y.position[out]), 
                xy.null = c(0, 0), color = col.out.synapse, fontsize = fontsize, 
                ...)
            grid.circle(x = x.position[out], y = y.position[out], 
                r = radius, gp = gpar(fill = "white", col = col.out, 
                  ...))
        }
        if (intercept) {
            for (k in 1:length(weights)) {
                y.tmp <- 0
                x.intercept <- (x.position[k + 1] - x.position[k]) * 
                  intercept.factor + x.position[k]
                for (i in 1:neuron.count[k + 1]) {
                  y.tmp <- y.tmp + y.step[k + 1]
                  result <- calculate.delta(c(x.intercept, x.position[k + 
                    1]), c(y.intercept, y.tmp), radius)
                  x <- c(x.intercept, x.position[k + 1] - result[1])
                  y <- c(y.intercept, y.tmp + result[2])
                  grid.lines(x = x, y = y, arrow = arrow(length = unit(0.15, 
                    "cm"), type = "closed"), gp = gpar(fill = col.intercept, 
                    col = col.intercept, ...))
                  xy.null <- cbind(x.position[k + 1] - x.intercept - 
                    2 * result[1], -(y.tmp - y.intercept + 2 * 
                    result[2]))
                  if (show.weights) 
                    draw.text(label = weights[[k]][1, neuron.count[k + 
                      1] - i + 1], x = c(x.intercept, x.position[k + 
                      1]), y = c(y.intercept, y.tmp), xy.null = xy.null, 
                      color = col.intercept, alignment = c("right", 
                        "bottom"), fontsize = fontsize - 2, ...)
                }
                grid.circle(x = x.intercept, y = y.intercept, 
                  r = radius, gp = gpar(fill = "white", col = col.intercept, 
                    ...))
                grid.text(1, x = x.intercept, y = y.intercept, 
                  gp = gpar(col = col.intercept, ...))
            }
        }
        if (information) 
            grid.text(paste("Error: ", round(result.matrix[rep, 
                "error"], 6), "   Steps: ", result.matrix[rep, 
                "steps"], sep = ""), x = 0.5, y = information.pos, 
                just = "bottom", gp = gpar(fontsize = fontsize + 
                  2, ...))
        popViewport()
        if (!is.null(file)) {
            weight.plot <- recordPlot()
            save(weight.plot, file = file)
        }
    }
}

calculate.delta <-
function (x, y, r) 
{
    delta.x <- x[2] - x[1]
    delta.y <- y[2] - y[1]
    x.null <- r/sqrt(delta.x^2 + delta.y^2) * delta.x
    if (y[1] < y[2]) 
        y.null <- -sqrt(r^2 - x.null^2)
    else if (y[1] > y[2]) 
        y.null <- sqrt(r^2 - x.null^2)
    else y.null <- 0
    c(x.null, y.null)
}

draw.text <-
function (label, x, y, xy.null = c(0, 0), color, alignment = c("left", 
    "bottom"), ...) 
{
    x.label <- x[1] + xy.null[1]
    y.label <- y[1] - xy.null[2]
    x.delta <- x[2] - x[1]
    y.delta <- y[2] - y[1]
    angle = atan(y.delta/x.delta) * (180/pi)
    if (angle < 0) 
        angle <- angle + 0
    else if (angle > 0) 
        angle <- angle - 0
    if (is.numeric(label)) 
        label <- round(label, 5)
    pushViewport(viewport(x = x.label, y = y.label, width = 0, 
        height = , angle = angle, name = "vp1", just = alignment))
    grid.text(label, x = 0, y = unit(0.75, "mm"), just = alignment, 
        gp = gpar(col = color, ...))
    popViewport()
}


plotNNpreds2D2class <- function(inpNN,inpClassThresh,inpGrid=(-60:60)/10) {
  
  tmpClrPch <- as.numeric(factor(inpNN$response))
  
  plot(inpNN$net.result[[1]],col=tmpClrPch,pch=tmpClrPch)
  
  table(inpNN$net.result[[1]][,1]>inpClassThresh,inpNN$response)
  
  xyGridTmp <- cbind(X1=rep(inpGrid,length(inpGrid)),X2=sort(rep(inpGrid,length(inpGrid))))
  
  # before predict.nn existed:
  #gridValsTmp <- compute(inpNN,xyGridTmp)$net.result
  gridValsTmp <- predict(inpNN,xyGridTmp)
  errTmp <- sum(inpNN$err.fct(inpNN$net.result[[1]][,1],inpNN$response))
  
  plot(xyGridTmp,
       col=as.numeric(gridValsTmp>inpClassThresh)+1,
       pch=20,cex=0.3,
       main=paste("Error:",round(errTmp,6)))
  points(inpNN$covariate,col=tmpClrPch,pch=tmpClrPch)
  
  ## Equations defining decision boundary:
  ## 1*w0 + X1*w1 + X2*w2 = 0, i.e.:
  ## 0 = inpNN$weights[[1]][1]+inpNN$weights[[1]][2]*X1+inpNN$weights[[1]][3]*X2, i.e:
  ## X2 = (-inpNN$weights[[1]][1] - inpNN$weights[[1]][2]*X1) / inpNN$weights[[1]][3]
  for ( iTmp in 1:ncol(inpNN$weights[[1]][[1]]) ) {
    abline(-inpNN$weights[[1]][[1]][1,iTmp] / inpNN$weights[[1]][[1]][3,iTmp],
           -inpNN$weights[[1]][[1]][2,iTmp] /inpNN$weights[[1]][[1]][3,iTmp],
           lty=2,lwd=2)
  }
}

```


# NN on 2D dataset: 2 centers 2 category side by side

## Build the dataset

```{r, fig.height=4,fig.width=5}
set.seed(1234321)

# Number of observations per category
n <- 1000

# Init a matrix with random samples
# Categories a and b will be placed in two different centers.
ab2d2c <- matrix(rnorm(2*n), ncol=2)

ctr_positions <- 1

# Randomly pull either a -1 or a +1
ab_ctrs_right <- matrix(sample(+1*ctr_positions, n, replace=TRUE), ncol=2)
ab_ctrs_left <- matrix(sample(-1*ctr_positions, n, replace=TRUE), ncol=2)

# Add to the ab random 2D variable to move the centers
ab2d2c[1:(n/2), 1] <- ab2d2c[1:(n/2), 1] + ctr_positions

beg <- n/2 + 1
ab2d2c[beg:n, 1] <-  ab2d2c[beg:n, 1] - ctr_positions

# categories
category <- vector(mode = 'numeric', 1000)
category[1:n/2] <- 'cat0' 
category[((n/2)+1):1000] <- 'cat1' 

# Plot with my colors
plot(ab2d2c ,col=as.numeric(factor(category)), pch=as.numeric(factor(category)), 
     xlab="V1", ylab="V2")
abline(h=0, col = 'red')
abline(v=0, col = 'red')
```

## Fit the model

```{r}
nn2d2c <- neuralnet(df ~ ., data.frame(df = as.numeric(factor(category)), ab2d2c))

```

## Plot the NN

```{r}
plot(nn2d2c)

```
## Predict

Model predictions recalculated from input variables from field 'covariate' and model field 'weights'. Needs activation function (fields `weights` and `act.fct`):

```{r}
head(nn2d2c$net.result[[1]])

```

```{r}
cbind(rep(1,6), nn2d2c$act.fct(cbind(rep(1,6), nn2d2c$covariate[1:6,]) %*% 
                              nn2d2c$weights[[1]][[1]]))%*%nn2d2c$weights[[1]][[2]]
```


# 2D dataset: 4 centers 2 category

## Build the dataset

```{r, fig.height=6,fig.width=6}
set.seed(1234321)

# Number of observations per category
n <- 1000

# Init a matrix with random samples
# Categories a and b will be placed in two different centers.
ab2d4c <- matrix(rnorm(4*n), ncol=2)

ctr_positions <- 2

# Randomly pull either a -1 or a +1
ab_centers <- matrix(sample(c(-1,1)*2, n*4, replace=TRUE), ncol=2)

# We get a mix os +/+, +/-, -/- -/+ times the number of observations
# Use them as locations for each of the sampled dataset observations.
ab2d4c <- ab2d4c + ab_centers

# Apply this formula to classify
# Based on script from Harvard Statistical Learnin by Dr. Sivachenko, Dr. Farutin
# It concatenates a zero or a 1 to the strig 'cat' for category
category <- paste0("cat", (1 + sign(apply(ab_centers, 1, prod)))/2)


# Plot it
plot(ab2d4c ,col=as.numeric(factor(category)), pch=as.numeric(factor(category)), 
     xlab="V1", ylab="V2")
abline(h=0, col = 'red')
abline(v=0, col = 'red')
```


## Fit the model

```{r}
nn2d4c <- neuralnet(df ~ ., data.frame(df = as.numeric(factor(category))-1, ab2d4c),
                    linear.output = FALSE, err.fct="ce")
plot(nn2d4c)
```



## Predict

Model predictions recalculated from input variables from field 'covariate' and model field 'weights'. Needs activation function (fields `weights` and `act.fct`):

```{r}
head(nn2d4c$net.result[[1]])
quantile(nn2d4c$net.result[[1]])
```

```{r}
cbind(rep(1,6), nn2d4c$act.fct(cbind(rep(1,6), nn2d4c$covariate[1:6,]) %*% 
                              nn2d4c$weights[[1]][[1]]))%*%nn2d4c$weights[[1]][[2]]
```
## Plot predictions

```{r}
old.par <- par(mfrow=c(1,2),ps=16)
plotNNpreds2D2class(nn2d4c,1.3)
par(old.par)

```

# 3D dataset: 1 center 2 category within our outside a cube

## Build the dataset

```{r 3d_1ctr_2cats, fig.height=5, fig.width=5}
##
#
# Call the function defined above to return our dataset.
# Then plot using scatterplot3d
# and insert a box based on help(scatterplot3d) function by Martin Maechler
#

# default values are n observations = 1000, x, y, z, limit is 1.25
set.seed(123321)

xyz <- bld_3d_2c()
class(xyz)
dim(xyz)
summary(xyz)
# Plot the cube

colors <- colors[as.numeric(factor(xyz[,1]))]

# http://www.sthda.com/english/wiki/scatterplot3d-3d-graphics-r-software-and-data-visualization
plot3d <- scatterplot3d(xyz[,2:4],
              color = colors,
              pch = xyz[,1])
cubedraw(plot3d, min = -1.25, max = 1.25, cex = 2)

```

## Fit the model: 1 hidden node

```{r}
df <- as.data.frame(xyz)
df$cl <- as.factor(df$cl)

nn3d1c_1h <- neuralnet(cl~., df, hidden=2, stepmax = 1e6,
                         linear.output = FALSE, err.fct = "sse")

# nn3d1c_12h <- neuralnet(cl~., df, hidden=c(1, 2), stepmax = 1e6,
#                          linear.output = FALSE, err.fct = "sse")
plot(nn3d1c_1h)

```
## Predict

```{r}
head(nn3d1c_1h$net.result[[1]])
quantile(nn3d1c_1h$net.result[[1]])
```

## System time example

```{r problem_2_sys_test}
#
# Since this is a classification problem, use linear.oputput = FALSE
#
df <- as.data.frame(xyz)
df$cl <- as.factor(df$cl)

system.time(invisible(neuralnet(cl~., df, hidden=2, stepmax = 1e6,
                         linear.output = FALSE, err.fct = "sse")))

system.time(invisible(neuralnet(cl~., df, hidden=c(1, 2), stepmax = 1e6,
                         linear.output = FALSE, err.fct = "sse")))

# Code with "ce" took too long
# system.time(invisible(neuralnet(cl~., xyz.df, hidden=3, stepmax = 1e6,
#                          linear.output = FALSE, err.fct = "ce")))
```

## Fit the model: c(1, 2) hidden node

```{r}
df <- as.data.frame(xyz)
df$cl <- as.factor(df$cl)

nn3d1c_12h <- neuralnet(cl~., df, hidden=c(1, 2), stepmax = 1e6,
                         linear.output = FALSE, err.fct = "sse")

plot(nn3d1c_12h)

```


## Predict

```{r}
head(nn3d1c_12h$net.result[[1]])
quantile(nn3d1c_12h$net.result[[1]])
```

