---
title: "R Intro"
author: 'Oscar A. Trevizo'
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    keep_md: yes
  github_document:
    toc: yes
---

```{r setup, include=TRUE}
# library(ISLR)
# library(RColorBrewer)
# library(reshape2)
# library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)

# The following tip shows how to force the creation of ONLY and md file.
# https://stackoverflow.com/questions/39814916/how-can-i-see-output-of-rmd-in-github

```

# Basics

Create chunck with Ctrl+Alt+I (Windows) Ctrl+Option+I (Mac)

```{r}

# Assign data to a variable

a <- 8

b <- 4

a + b

# Patterned data
2:8
1:7



```

# Data types

* Vector, 
* Matrix, 
* Array, 
* List, 
* Data Frame

## Vector

```{r}
##
#
# Vector
#
# Atomic types:
# logical, integer, numeric, complex, character
#
# All the values must have a consistent data type within a vector
#
##

vtr_logical = c(TRUE, TRUE, FALSE, FALSE, TRUE)
vtr_logical

vtr_integer = c(256L, 1024L, 16L)
vtr_integer

vtr_numeric = c(2.718, 3.1416, 0.7071)
vtr_numeric

vtr_char = c("hello, world", "now i'm here", "now i'm there")
vtr_char

```

## Operations on vectors

```{r}
##
#
# Let's play with vectors (it is like NumPy arrays)
#
##

a <- c(1, 2, 3)

a + 1

a / 2

a * 2

b <- c(4,5,6)

a + b

a - b


```
## Plot vectors

The plot function takes two vectors, one for X values and one for Y values, and draws a graph of them


```{r}

x <- seq(1, 20, 0.1)
y <- sin(x)

plot(x, y)

```

## Another plot example

```{r}

values <- -10:10
absolutes <- abs(values)

plot(values, absolutes)

```

## NA values

```{r}

a <- c(1, 3, NA, 7, 9)

# Here we will get a NA
sum(a)

# Try calling sum again, with na.rm set to TRUE:
sum(a, na.rm=TRUE)

```
# Matrix

```{r}

#
# Matrix
#

mtx1 = matrix(c(1:25), 5, 5)
mtx1

mtx2 = matrix(c(1:6), 2, 3)
mtx2

```

## Array

```{r}
#
# Array
#

# This is an array of matrices

# Six values (1:6), then 2 rows x 3 columns matrices, 
# Then each matrix is identified by x, y (4x2), that is 8 of them
arr1 = array(c(1:6), dim=c(2, 3, 4, 2))
arr1

```

# Lists

```{r}
#
# Lists
#

list1 = list(vtr_char, vtr_integer, vtr_logical, vtr_numeric)
list1

```

## Data Frame

```{r}
#
# Data Frame
#

vtr_char_names = c("Jay", "Julie", "John")

# Some vectors define in previos cells
data.frame(vtr_char_names, vtr_integer, vtr_numeric, vtr_char)

# Another dataframe
grades <- c(90, 85, 92, 75, 88)
students <- c('Joe', 'Mary', 'Gina', 'Vijay', 'Jay')

st_data <- data.frame(students, grades)

# To get the stucture of the data, the metadata
str(st_data)


```

# Built-in datasets

```{r}

data()
head(iris)
iris[2:4, 4:5]

is = iris

summary(is)

# Standard deviation of a particular column
sd(iris$Sepal.Length)

# airquality is a table already built in. let's use it
aq = data.frame(airquality)

```

# Vectorization / time stamps

```{r vectperf}
##
#
# See how long it takes to do something
#
##

# Create some data

# Assign a matrix to variable x. Fill with normally distr random numbers
# Make it 10 columns by 50000 rows
x <- matrix(rnorm(50000*10), ncol=10)

# Make y numeric (coerse)
y <- numeric()

# Time stamp
pt1 <- proc.time()

# Iterate
for ( i in 1:dim(x)[1] ) y[i] <- mean(x[i,])

# Time stamp and then calculate how long it tok between pt1 and pt2
pt2 <- proc.time(); pt2-pt1; y[1:3]

# apply() is like Python lambda. Here, calculate the mean
y <- apply(x, 1, mean)

# Stamp time again, and measure the difference
pt3 <- proc.time(); pt3 - pt2; y[1:3]

y <- rowMeans(x)
proc.time() - pt3; y[1:3]
```

# Operators

```{r}
##
#
# Operators
#
# Arithmetic, Assignment, Relational, Logical
#

# Artithmetic
#
cat('3+5.5 = ', 3+5.5, '\n')
print(15/3)
print(2^7)
print(22/7)

# modular division
print(22%%7)

# flow division rounds up to previous whole number
print(22%/%7)

# Relational opertors
# (compares)
#

var1 = 5
var2 = 12
print(var1 > var2) #the result is FALSE
print(var1 == var2) #the result is FALSE
print(var1 != var2) #the result is TRUE
print(var1 < var2) #the result is TRUE

# Assignment operators = or <- or -> left or right either way
#
# examples of assigning a value to x
#

x <- 15
x
x <- 4
x
x = 8
x
25 -> x
x

# Logical operators
#
# &, |, !

vtr2_logical = c(TRUE, FALSE,TRUE, FALSE, TRUE)
vtr3_logical = c(TRUE, TRUE, TRUE, TRUE, TRUE)
vtr4_logical = c(FALSE, FALSE, FALSE, FALSE, FALSE)

print(vtr2_logical & vtr3_logical)
print(vtr2_logical | vtr3_logical)
print(!vtr2_logical)
print(vtr2_logical && vtr3_logical)
print(vtr2_logical || vtr3_logical)


```

# Conditional statements

```{r}
##
#
# Conditional Statemets
#
##

#
# if, else if
#

x = 3
if(x==5)
{
  print("x is equal to 5")
} else if(x > 5)
{
  print("x is greater than 5")
} else if(x<5)
{
  print("x is smaller than 5")  
}


#
# Switch Case Statement
#

x = 4   # it will look for the xth item
switch(x,
'1' = print("It's one"),
'2' = print("It's two"),
'3' = print("It's three"),
'4' = print("It's four"),
'5' = print("It's five"),
'6' = print("It's six"),
'7' = print("It's seven"),
'8' = print("It's eight"),
'9' = print("It's nine"),
'?' = print("I don't know, what is it?")
)

```


# Repeat Loop

```{r}
# repeat{}

i = 0
repeat
{
  print(i)
  i = i + 1
  if(i > 9)
  {
    break
  }
}

```


# While Loop

```{r}

i = 0
while(i < 10)
{
  print(i)
  i = i + 1
}

```


# For Loop

Run over an iterator

## Example 1: Basic for loop

```{r} 

for (i in 0:9)
{
  print(i)
}


```


## Example 2: Loop a normal distribution

Reference: Dr. Bharatendra https://www.youtube.com/watch?v=_hHQZP7_52Y&list=PL34t5iLfZddtUUABMikey6NtL05hPAp42

```{r}

N <- 4

# Plot control, 2 x 2
par(mfrow=c(2, 2))

for (i in 1:N){
  x <- rnorm(10)
  qqnorm(x)
  qqline(x)
}

```
## Example 3:

Reference: Dr. Bharatendra https://www.youtube.com/watch?v=_hHQZP7_52Y&list=PL34t5iLfZddtUUABMikey6NtL05hPAp42


```{r}
# Generate 50 means
M <- 50
# Coerce object M to type numeric
AVG <- numeric()

# Plot control, 1 x 1
par(mfrow=c(1, 1))

for (i in 1:M) {
  AVG[i] <- mean(rnorm(30))
}

# Display it
AVG

# Show histogram
hist(AVG)
```



# Strings

```{r}
##
#
# String
#
##

str1 <- "How're you doing?"
print(str1)
nchar(str1)

str2 <- "Doing fine."
print(str2)
nchar(str2)

str3 = paste(str1, str2)
print(str3)
nchar(str3)


```

# Functions

```{r}
##
#
# Functions
#
# Predefined and User Define
#
##

ftn1 <- function(x)
{
  x2 = x^2
  print(x)
  print(x2)
}

ftn1(3)


```

# Visualization

```{r}

##
#
# Visualization
#
# Pie Charts, Bar Chart, Boxplot, Histogram, Line Graph, Scatterplot
#
##

pie(vtr_numeric, vtr_char_names)   # very plain piechart

pie(vtr_numeric, vtr_char_names, col = rainbow(length(vtr_numeric)))

barplot(vtr_numeric)

# mtcars is a built in dataset

boxplot(mpg ~ cyl, data = mtcars, xlab = "Number of Cylinders", 
        ylab = "Miles per Galon", main = "Milage Data",
        col=c("red", "yellow", "blue"))

vtr5_numberic = c(1,3,5,2,8,8,5,6,6,5,2,9,2,9,8)

hist(vtr5_numberic)

length(vtr5_numberic)
vtr6_numberic = c(1:length(vtr5_numberic))
vtr6_numberic

plot(vtr6_numberic, vtr5_numberic)

```

## pair plots

```{r}

is = data(iris)

pairs(iris[,c(1:4)])
```

```{r}

```

# Load data -- Read CSV

```{r}

df <- read.csv("../data/fifa_cleaned.csv")

```


