---
title: "Build Dataset: Human migration dataset based on UN WPP and World Bank"
author: "Oscar A. Trevizo"
date: "2023-05-08"
output:
  pdf_document:
    toc: yes
    number_sections: yes
    toc_depth: 4
  html_document: 
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

At the end of this use case, we will have a dataset built on sources from the United Nations and the World Bank to provide relationships between migration, demographics, socioeconomic, and development indicators.

# Load the libraries


```{r libraries}
# Essential libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(psych)
library(GGally)
library(PerformanceAnalytics)



```

# United Nations WPP

## Load the data


```{r read_un_wpp}
# https://population.un.org/wpp/Download/Files/1_Indicators%20(Standard)/\
# EXCEL_FILES/1_General/WPP2022_GEN_F01_DEMOGRAPHIC_INDICATORS_REV1.xlsx
# wpp_raw <- read_excel('data/wpp.xlsx',
#                  sheet = 'Estimates',
#                  skip = 16,
#                  col_types = 'text')

# Saved as CSV
wpp_raw <- read.csv('../data/WPP2022_GEN_F01_DEMOGRAPHIC_INDICATORS_REV1.csv', check.names=FALSE)
```


## Create WPP data set

```{r wpp_filter}
# The first 3 columns listed below are only informative.
# Total population is redundant. Keeping the July data.
# Notice, Year is class 'int'
wpp <- wpp_raw %>% filter((Type == 'Country/Area')) %>% 
  dplyr::select(-c('Variant',
            'Notes',
            'Location code',
            'SDMX code**',
            'Total Population, as of 1 January (thousands)'))

dim(wpp)
head(wpp)
```

## Add Columns Regions and Subregions

```{r}
# Get Subregions in a small table.... One year takes care of having one row per region.
# We need to include both Subregion and Region because the USA, Canada, Greenland,... 
# have NORTH AMERICA as Parent Code while the other countries have Subregion as
# Parent Code. Pick one year, any year like 2019 to create these temporary dataframe.
wpp_parent_code <- wpp_raw %>% filter((Type == 'Subregion' | Type == 'Region') 
                                     & Year == 2019) %>% 
  dplyr::select(c('Region, subregion, country or area *',
                  'Location code'))

# Rename columns in preparation for a merge that will be done based on 'Parent code'
colnames(wpp_parent_code) <- c('Subregion', 'Parent code')

# Now merge with wpp based on 'Parent code'
wpp <- wpp %>% left_join(wpp_parent_code)

# Now we have subregions.
# The next step is to add Region (COntinent / subcontinent). The problem there
# is NORTH AMERICA is Region but also Parent Code for USA and Canada.

wpp_subregions <- wpp_raw %>% filter(Type == 'Subregion'
                                     & Year == 2019) %>% 
  dplyr::select(c('Region, subregion, country or area *',
                  'Parent code'))

wpp_regions <- wpp_raw %>% filter(Type == 'Region'
                                     & Year == 2019) %>% 
  dplyr::select(c('Region, subregion, country or area *',
                  'Location code'))

# Rename columns in wpp_regions to match the column name 'Parent code'
# before left_join(). this way the function automatically will pick Parent code

colnames(wpp_regions) <- c('Region', 'Parent code')
colnames(wpp_subregions) <- c('Subregion', 'Parent code')

# now the join will automaticaly use 'Parent code' to join left.
wpp_subregions <- wpp_subregions %>% left_join(wpp_regions)

# Remove columns 'Parent Code' no longer needed
wpp_subregions <- wpp_subregions %>% dplyr::select(-c('Parent code'))

# But we need to add NORTH AMERICA... Let's follow the Upper Case Lower Case approach
wpp_subregions <- rbind(wpp_subregions, c('NORTHERN AMERICA', 'NORTHERN AMERICA'))


# Now merge with wpp based on 'Subregion'
wpp <- wpp %>% left_join(wpp_subregions)


```


## Rename and re-sort columns


### Save column names and create a table


```{r wpp_new_column_names}
wpp_columns_orig <- colnames(wpp)
wpp_columns_new <- c('index', 'country', 'ISO3', 'ISO2', 'type', 'parent_code', 'year',
                     'pop', 'pop_m', 'pop_f', 'pop_density', 'pop_sex_ratio', 'med_age',
                     'natural_change', 'natural_change_rate', 'pop_change',
                     'pop_growth_rate', 'pop_doubling', 'births', 'births_w15to19',
                     'birth_rate', 'fertility_rate', 'net_reproduction_rate',
                     'mean_age_childbearing', 'sex_ratio_birth', 'tot_deaths',
                     'male_deaths', 'female_deaths', 'death_rate',
                     'life_exp', 'life_exp_m', 'life_exp_f',
                     'life_exp_15', 'life_exp_15_m', 'life_exp_15_f',
                     'life_exp_65', 'life_exp_65_m', 'life_exp_65_f',
                     'life_exp_80', 'life_exp_80_m', 'life_exp_80_f',
                     'infant_deaths', 'under_five_mortality',
                     'live_births', 'deaths_under_5', 'mortality_rate_under_5',
                     'mortality_40', 'mortality_40_m', 'mortality_40_f',
                     'mortality_60', 'mortality_60_m', 'mortality_60_f',
                     'mortality_15_50', 'mortality_15_50_m', 'mortality_15_50_f',
                     'mortality_15_60', 'mortality_15_60_m', 'mortality_15_60_f',
                      'net_migrants', 'net_mig_rate', 'subregion', 'region'
                     )
```

### Column definitions

```{r}
wpp_columns_table <- data.frame(wpp_columns_orig, wpp_columns_new)
wpp_columns_table
```


### Rename wpp columns

```{r}
# Rename dataframe columns
(colnames(wpp) <- wpp_columns_new)
```

### Re-sort the columns

```{r}

# Obtain numeric columns
col_numeric <- wpp_columns_new[7:60]

# Obtain non-numeric, character, columns... It is a handful, sort it here by hand
col_char <-  c('index', 'country', 'subregion', 'region', 
               'ISO3', 'ISO2', 'type', 'parent_code')

wpp <- wpp %>% dplyr::select(col_char, col_numeric)

colnames(wpp)
```

## Column variable types

Now we need to make numeric columns be type numeric.

```{r}
# Remove white spaces from numeric columns
# https://www.geeksforgeeks.org/remove-all-whitespace-in-each-dataframe-column-in-r/
wpp[, 9:62] <- as.data.frame(apply(wpp[, 9:62], 2, function(x) gsub("\\s+", "", x)))

# Coerce numeric
# Tips from:
# https://stackoverflow.com/questions/2288485/how-to-convert-a-data-frame-column-to-numeric-type
wpp[, 9:62] <- sapply(wpp[, 9:62], as.numeric)

# Year should actually be Date as this:
# The challenge with a Date is it will try to pin down the exact month and day too.
# Even if the format is Year only, underneath the data will have exact days
# wpp$Year <- as.Date(wpp$year, '%Y')

# But for now, leave it as numeric for simplicity.
wpp$year <- as.numeric(wpp$year)

# Remove columns 'type' and 'parent_code' no longer needed
wpp <- wpp %>% dplyr::select(-c('type', 'parent_code'))

```


## Remove some columns


```{r}

# Remove pop_doubling and live_births columns, which have NAs
wpp <- wpp %>% dplyr::select(-c('pop_doubling', 'live_births'))

```

## Add columns

### Add migration stat variables

Describe stats for migration

```{r}
describe(wpp[,c("net_migrants", "net_mig_rate")])
```



### Add columns with stats on migration


```{r}
# Use dplyr to find mu, sigma, and quantiles

wpp_mig_stats <- wpp %>% group_by(ISO3) %>% 
  summarise(net_mig_rate_mean = mean(net_mig_rate, na.rm = TRUE),
            net_mig_rate_sd = sd(net_mig_rate, na.rm = TRUE),
            net_mig_rate_med = median(net_mig_rate, na.rm = TRUE),
            net_mig_rate_25qtl = quantile(net_mig_rate, probs = 0.25, na.rm = TRUE),
            net_mig_rate_75qtl = quantile(net_mig_rate, probs = 0.75, na.rm = TRUE)
            )

# Now join those values to the data set.
wpp <- wpp %>% left_join(wpp_mig_stats)
```


### Create factor variables for migration

#### Create new factor variable: emigrates (1, 0)

This new column will have a 1 if the country emigrates (leaves), and a 0 where they
don't migrate in net figures.

Call it emigrates. 
Gets a 1 if it has a negative migration. That is, more people emigrate (like Mexico).
Gets a 0 if it has a positive or zero migration. 
That is, they experience immigration (like the US).

That is, if it has a negative net_migration, the country emigrates (leaves), and if it 
has a positive net_migration, the country receives migrants.

#### New columns migration pattern mig_pattern: factor 'emigration' or 'immigration'

Countries may experience a mix of years when they have emigration (leaving) and other
years when they have immigration. For the most part the pattern is consistent.
Some countries consistently receive immigrants (like the United States and Germany) 
and other countries consistently have emigrants leave the country.

To avoid a possible mix between emigration and immigration within a country, this 
variable will classify the country based on its general pattern. The variable
will be class as factor and will have the values:

- emigration
- immigration

To add this column we need to go through a two step process.

```{r}
# First remove any country that has NA in net_mig_rate.
# Upon exploration, there is only one country as such = VAT (the Vatican)
# It is best to remove it as it does not show any other patterns.
# (note, there was another command before that removed the Vatican)
# (leave this one here for now, even though it is is not doing anything)
wpp <- wpp %>% drop_na(net_mig_rate)

# This command create the new column 'emigrates' with values either 1 or 0.
# Note that the same country may have a mix of 1s and 0s in different years.
# This variable can change from 1 to 0 or 0 1 for a country from year to year
wpp <- wpp %>% mutate(emigrates = ifelse(net_mig_rate < 0, 1, 0))

# Mig patterns is a category. We want to assign a category to the country based on an
# overall migration pattern over the years. 
# Emigration means the country tends to have emigrants (a negative mig rate overall),
# while immigration means the country tends to receive migrants (a positive mig rate).
# We will leave 0 as immigration to avoid having three categories.
# But there was no single country with an exact 0.0000 mig rate mean.
mig_pattern_df <- wpp %>% group_by(ISO3) %>%
  summarise(mig_pattern = ifelse(median(net_mig_rate, na.rm = TRUE) > 0,
                                 'immigration', 'emigration'))

# Now join that pattern grouped by ISO3 (country) with the overall data set wpp
wpp <- wpp %>% left_join(mig_pattern_df)


```



#### Create factor variable for migration_swings 

Some countries exhibit swings, seasonal patterns (e.g. Qatar), while other countries
exhibit a controlled migration pattern. That question can be explored by adding three
additional columns to our data set, focusing on migration rate:

If a country exhibits swings between the 25 and 75 quantiles from negative to positive
migration, the country needs to be identified as one that exhibits migration
swings, Those countries could have seasonal migration, leaving the country and 
then entering the country again. That is a special category.


```{r}
wpp <- wpp %>% mutate(migration_swings = ifelse(net_mig_rate_75qtl * net_mig_rate_25qtl < 0, 1, 0))

```



```{r}
wpp_mig_stats_swings <- wpp_mig_stats %>% 
  mutate(migration_swings = ifelse(net_mig_rate_75qtl * net_mig_rate_25qtl < 0, 1, 0))

table(wpp_mig_stats_swings$migration_swings)
```



# World Bank

https://www.worldbank.org/en/home

## DataBank

https://databank.worldbank.org/

- World Development Indicators
- Statistical Capacity Indicators
- Education Statistics
- Gender Statistics
- Health Nutrition and Population Statistics


## World Development Indicators

Tool to select multitude variables to characterize countries or regions. It allows us
to create custom report. The reports can be shared or can be private.

There are $1442$ _Series_ (i.e., variables, predictors) available. 



## WorldBank databank report

The tool allows you to select a layout, save the report and share the report. I created a shared report from WorldBank databank:

https://databank.worldbank.org/Countries-GDP-and-Inflation/id/4f7a6fed

## Load the data

```{r}
# From WOrld bankd 0 Databank:
# https://databank.worldbank.org/id/49085302?Report_Name=Countries-GDP-Inflation-and-predictors&l=en
# https://databank.worldbank.org/Countries-GDP-Inflation-and-predictors/id/18e02f13
# Saved as CSV for convenience.
#
# wb_raw <- read.csv('data/wb.csv', check.names=FALSE, colClasses = 'character')
wb_raw <- read.csv('../data/wb_2023.Q1.csv', check.names=FALSE)

wb_metadata <- read.csv('../data/wb_metadata_2023.Q1.csv', check.names=FALSE, colClasses = 'character')

```


## World Bank data set definitions

- From WB metadata

```{r}
wb_metadata[,c('Indicator Name', 'Long definition')]
```

## Create WPP data set


```{r}
# Use dplyr select() for simplicity. Put it in a new dataframe wb.
# Notice Time is class 'int'... That is how read.csv pulled it in. 
wb <- wb_raw %>%  dplyr::select(-c('Country Name', 'Time Code', 
                                  'International migrant stock (% of population) [SM.POP.TOTL.ZS]',
                                  'Net migration [SM.POP.NETM]'))
```


## Rename columns

Anticipating a merge between the data sets from the World Bank (economics) with
the United Nations (demographics), rename variables in the _wb_ data set to match
columns in the _wpp_ data set. Specifically:

- ISO3: Country code
- Year: The time period.

Simultaneously, rename the remaining columns to something more practical
from the scripting perspective.

Save the old names and the new names in a new table for the record.

```{r}
wb_columns_orig <- colnames(wb)
wb_columns_new <- c('ISO3', 
                    'year', 
                    'GDP', 
                    'GDP_growth', 
                    'GDP_pc', 
                    'GDP_pc_ppp',
                    'inflation',
                    'natl_income_pc',
                    'cpi',
                    'food',
                    'homicides',
                    'electricity',
                    'co2_emissions',
                    'education',
                    'mig_stock'
                    )
```


### Column definitions

```{r}
wb_columns_table <- data.frame(wb_columns_orig, wb_columns_new)

wb_columns_table
```

### Rename columns

```{r}
(colnames(wb) <- wb_columns_new)
# head(wb)
```

## Replace '..' values with NA  

```{r}
# Replace across the entire data set, all columns, specifically applies to numerics.
wb[wb == '..'] <- NA
head(wb)
```

## Coerce chars as numeric

```{r}
# Remove white spaces from numeric columns
# https://www.geeksforgeeks.org/remove-all-whitespace-in-each-dataframe-column-in-r/
wb[, 3:15] <- as.data.frame(apply(wb[, 3:15], 2, function(x) gsub("\\s+", "", x)))

wb[, 3:15] <- sapply(wb[, 3:15], as.numeric)

# Year should actually be Date:
# The challenge with a Date is it will try to pin down the exact month and day too.
# Even if the format is Year only, underneath the data will have exact days
# wb$year <- as.Date(wb$year, '%Y')

# But for now, leave it as numeric for simplicity.
wb$year <- as.numeric(wb$year)
```



## WB Summary

```{r}
# It's helpful to see the min and max in determining which variables are candidates for log transformation.
summary(wb)
```

## Correlation pairs

### Economics indicators

```{r}
wb[,3:9] %>% pairs.panels()

```

### Development indicators

```{r}
wb[,11:15] %>% pairs.panels()

```

## Transforms

Candidates for $log()$ transformation are all of the economic variables except perhaps for _GDP_growth_:

- GDP
- GDP_pc
- GDP_pc_ppp
- inflation
- natl_income_pc
- cpi
- mig_stock


```{r}
# Instead of replacing, add new log variables using dplyr mutate()
wb <- wb %>% mutate(log_GDP = log(GDP),
                    log_GDP_pc = log(GDP_pc),
                    log_GDP_pc_ppp = log(GDP_pc_ppp),
                    log_inflation = log(inflation + 20),
                    log_natl_income_pc = log(natl_income_pc + 500),
                    log_cpi = log(cpi),
                    log_mig_stock = log(mig_stock))
head(wb)
```

## Summary and correlation pairs of log transformed variables


```{r}
summary(wb[16:22])
wb[16:22] %>% pairs.panels()

```


Do $log()$ transformation also for:

- food
- homicides
- electricity
- co2_emissions
- education

For simplicity, _log_ transform all the numeric variables.


```{r}
# Instead of replacing, add new log variables using dplyr mutate()

wb <- wb %>% mutate(log_food = log(food),
                    log_homicides = log(homicides + 0.01),
                    log_electricity = log(electricity),
                    log_co2_emissions = log(co2_emissions),
                    log_education = log(education + 0.01))
head(wb)
```

```{r}
summary(wb[,23:27])
wb[23:27] %>% pairs.panels()

```


# Build the dataframe

## Join WPP with WB by ISO3 and Year

We will use _dplyr_ to perform the join, a.k.a. _merge_ of two data sets.

We can either do a _left, inner,_ or a _right_ join.  The order depends on which 
data set we pass first as an argument. 

On a _left_ join, the data set passed first (left argument) serves as the main data set. 
Matching records are added as new columns from the second data set. As a result, the number
of rows will be determined by the _left_.

On an _inner_ join, the two data sets are matched. The result will consists only of 
those observations/rows that are found in both data sets. Therefore, the number of rows 
will be less than or equal to that of the smaller data set.

And on a _right_ join, the second data set takes precedence.

Here we will use the UN WPP data set as the main data set. We will apply a _left_ join.

```{r}
wpp_wb <- wpp %>% left_join(wb)
```

Notice the resulting data set _wpp_wb_ has the same number of observations as _wpp_.

Now we have a data set with demographics and with economics data.


## Reorder the data set columns

- Place non-numeric variables first
- Then outcome variables (migration related)
- Then the rest of the numeric variables.

## Column names wpp_wb

```{r}
colnames(wpp_wb)
```

## Separate non-numeric from outcome and predictor variables

```{r}
# First group
# These describe/identify the observations, e.g. country/region name, ISO, year
non_num_cnames <- colnames(wpp_wb[, 1:7])

# Second group, outcomes and related migration vars
# This step will remove some columns as well
outcomes <- c('net_mig_rate',
              'net_migrants',
              'mig_stock',
              'log_mig_stock',
              'emigrates',
              'mig_pattern',
              'migration_swings',
              'net_mig_rate_mean',
              'net_mig_rate_med')

# Third group - Predictors
num_cnames <- colnames(wpp_wb[, 8:91])

predictors <- num_cnames[! num_cnames %in% c('net_mig_rate',
                                      'net_migrants',
                                      'migration_stock',
                                      'log_mig_stock',
                                      'emigrates',
                                      'mig_pattern',
                                      'migration_swings',
                                      'net_mig_rate_mean',
                                      'net_mig_rate_sd',
                                      'net_mig_rate_med',
                                      'net_mig_rate_25qtl',
                                      'net_mig_rate_75qtl'
                                )]

```


## Rearrange migration variables to the left of the numeric dataframe


```{r}

wpp_wb <- wpp_wb %>% dplyr::select(non_num_cnames, outcomes, predictors)

# make categorical variables as factors
wpp_wb$mig_pattern <- as.factor(wpp_wb$mig_pattern)
wpp_wb$emigrates <- as.factor(wpp_wb$emigrates)
wpp_wb$migration_swings <- as.factor(wpp_wb$migration_swings)

colnames(wpp_wb)
```

# Train test split

Build a train test split up-front to be used consistently throughout this analysis.

```{r}
# We have 236 countries.
countries <- unique(wpp_wb$country)

# Now get a sample of say 70% of that list
set.seed(12321) 
countries_sample <- sample(countries, length(countries)*0.7)

# Now we will want train to contain those countries
# And test not to contain those countries.
train <- wpp_wb %>% filter(country %in% countries_sample)
test <- wpp_wb %>% filter(!country %in% countries_sample)

# Traditionally we would split the df up and down as follows
# set.seed(12321) 
# ind <- sample(2, nrow(wpp_wb), replace = T, prob = c(0.7, 0.3))
# train <- wpp_wb[ind == 1,]
# test <- wpp_wb[ind == 2,]

# Another option data scientists use when splitting data that originates
# from a time series is by training from year 1 to year x 
# and testing from years x+1 to last year. 


```

## Explore train and test sets

This shows the count of observations/rows:

### By region

```{r}
train %>% 
  group_by(region) %>% 
  summarize(count = n())

test %>% 
  group_by(region) %>% 
  summarize(count = n())

```


### By subregion

```{r}
train %>% 
  group_by(subregion) %>% 
  summarize(count = n())

test %>% 
  group_by(subregion) %>% 
  summarize(count = n())

```

### By country

```{r}
train %>% 
  group_by(country) %>% 
  summarize(count = n())

test %>% 
  group_by(country) %>% 
  summarize(count = n())

```

### By year

```{r}
train %>% 
  group_by(year) %>% 
  summarize(count = n())

test %>% 
  group_by(year) %>% 
  summarize(count = n())

```

The count of countries can be taken by looking at one single year (e.g. 2019)

### Countries by region

```{r}
train %>% filter(year == 2019) %>% 
  group_by(region) %>% 
  summarize(count = n())

test %>% filter(year == 2019) %>% 
  group_by(region) %>% 
  summarize(count = n())

```

# Write to CSV to be used downstream


```{r}
write.csv(wpp_wb, '../data/wpp_wb_2023.Q1.csv', row.names = FALSE)
write.csv(train, '../data/wpp_wb_train_2023.Q1.csv', row.names = FALSE)
write.csv(test, '../data/wpp_wb_test_2023.Q1.csv', row.names = FALSE)

```

# References

1.	United Nations Datamart - World Population Prospects (accessed Mar. 15, 2023) https://population.un.org/wpp/Download/Standard/MostUsed/
2.	World Bank World Development Indicators (accessed Mar. 16, 2022) https://databank.worldbank.org/source/world-development-indicators 
3.	World Bank World Development Indicators. Data_Extract (accessed Mar. 18, 2023) https://databank.worldbank.org/Countries-GDP-and-Inflation/id/4f7a6fed   
4.	Migration Policy Institute. “Immigration Has Been a Defining, Often Contentious, Element Throughout U.S. History.” Jan. 6, 2022 (accessed Mar 30, 2023) https://www.migrationpolicy.org/article/immigration-shaped-united-states-history 
5.   International Organization for Migration (IOM). "World Migration Report 2022." ISBN 978-92-9268-078-7. (accessed Apr. 24, 2023)  https://publications.iom.int/books/world-migration-report-2022  
6.	 International Organization for Migration (IOM) Global Migration Analysis Centre. Migration Data Portal (accessed Apr. 14, 2023) https://www.migrationdataportal.org/ 
7.	IOM Migration Research Series. “Gallup World Poll: The Many Faces of Global Migration.” https://news.gallup.com/poll/152660/Faces-Global-Migration.aspx 
8.	World Migration Report 2022. (accessed Apr. 10m 2023) https://migrationnetwork.un.org/resources/world-migration-report-2022

**Useful links**

- United Nations Datamart: http://data.un.org/
- United Nations Datamart - World Population Prospects (WPP): https://population.un.org/wpp/Download/Standard/MostUsed/
- United Nations Datamart - World Population Prospects Excel file: https://population.un.org/wpp/Download/Files/1_Indicators%20(Standard)/EXCEL_FILES/1_General/WPP2022_GEN_F01_DEMOGRAPHIC_INDICATORS_REV1.xlsx
- United Nations Datamart - Economic Trading: https://population.un.org/wpp/Download/SpecialAggregates/EconomicTrading/
- United Nations GDP: http://data.un.org/_Docs/SYB/CSV/SYB65_230_202209_GDP%20and%20GDP%20Per%20Capita.csv
- World Bank DataBank. Global Bilateral Migration:  https://databank.worldbank.org/source/global-bilateral-migration/Type/TABLE/preview/on#
- World Bank Inflation: https://data.worldbank.org/indicator/FP.CPI.TOTL.ZG?view=charthttps://data.worldbank.org/indicator/FP.CPI.TOTL.ZG?view=chart
- World Bank World Development Indicators P_Data_Extract_From_World_Development_Indicators https://databank.worldbank.org/source/world-development-indicators
- World Bank World GDP:  https://data.worldbank.org/indicator/NY.GDP.MKTP.KD.ZG file API_NY.GDP.MKTP.KD.ZG_DS2_en_excel_v2_4701280.xls
- US DHS Satistics: https://www.dhs.gov/immigration-statistics/naturalizations
- U.S. Naturalizations 2021 Data Tables: https://www.dhs.gov/sites/default/files/2022-09/2022_0624_plcy_naturalizations_fy2021_tables.zip
- US Naturalizations: https://www.dhs.gov/sites/default/files/2022-07/2022_0624_plcy_naturalizations_fy2021.pdf

